Calibrating seed-based mapping heuristics with Sesame

### 2020-03-22 manuscript.tex
Add background on spaced seeds

\title{Calibrating seed-based alignment heuristics with Sesame}

@@

@@

The most popular heuristic for mapping DNA sequences is a filtration
method called seeding. The principle is to first identify seeds,
defined as short regions of high similarity between the read and the
genome, and then use an exact alignment algorithm at the candidate
locations to identify the optimum. This strategy was first proposed in
FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to to search for
homology in sequence databases.

@@

The rest of this document is organized so as to present the concepts by
increasing level of complexity, meaning that we sometimes go back and
forth between different types of seeding strategies. We first provide some
background knowledge on different types of seeding methods, we then
introduce the computational strategy based on transfer matrices, and
finally we delve into each specific seeding method.

\section{Seeds}

The term ``seed'' has different meanings in computational biology. It can
designate a part of the read, a part of the genome, a particular sequence
motif, or a structured pattern of matches. Also, a seed does not always
refer to an exact match. For instance, the algorithm
PatternHunter~\cite{pmid11934743} uses ``spaced seeds'' that tolerate
mismatches. To avoid any confusion, we will adopt the convention below.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}. Hash
tables are typically used to index $k$-mers, which makes them useful to
search for seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
structures have no set size so they can be used to search for seeds of
different lengths. For instance, the FM-index is a compact data structure
based on the suffix array~\cite{manber1993suffix} and on the
Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
with a much smaller memory footprint~\cite{ferragina2000opportunistic,
ferragina2005indexing}.

@@

@@

\begin{definition}
A Maximal Exact Match (MEM) is a subsequence of the read that is present
in the reference genome and that cannot be extended --- either because the
read ends or because the extended subsequence is not in the genome.
\end{definition}

@@

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Figure~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
the target and the first duplicate (provided $\gamma \leq 12$).

@@

We need to know whether a nucleotide is a sequencing error, but this time
we also need to know its phase in the repeated cycles of skipped
positions. For this, we define the skip-$n$ alphabet $\mathcal{A}_n =
\{\square, |, \Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$. Again,
the symbol $\square$ represents a correct nucleotide and the symbol $|$ is
a terminator added at the end of the read. The symbols $\Downarrow_j$ ($ 0
\leq j \leq n$) represent sequencing errors and $j$ indicates the number
of nucleotides till the next non-skipped position (\textit{i.e.} $j=0$ for
nucleotides immediately before a non-skipped position and $j=n$ for
nucleotides at a non-skipped position).

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the $\square$ symbol
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$j$ in $\Downarrow_j$ indicates the number of nucleotides till the next
non-skipped position. For $j=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

Those segments follow a $\Downarrow_i$-segment, so they start $i$
nucleotides in front of the next non-skipped position. The shortest
following segment that ends $j$ nucleotides before a non-skipped position
has length $\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is
equivalent to defining $\ell_0$ as $x+1$ where $x$ is the number of
$\square$ symbols of the shortest $\Downarrow_j$-segment, \textit{i.e.} $x
= i-j-1 \modulo{n+1}$.

@@

@@

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, the top right entry of
$\tilde{M}_0(z)^{s+1}$ is the weighted generating function of reads that
contain $s$ mismatches (against any of the two sequences). We thus define
$\tilde{p}$ as the upper bound on the probability of a mismatch,
\textit{i.e.} $\tilde{p} = \max\{b,c,d\}$. The updated formula
(\ref{eq:bound}) now gives an upper bound of the probability that a read
of size $k$ contains $s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
be exactly identical to the target, making it impossible that there is a
seed for a duplicate but no seed for the target. When $\mu$ is much
larger, the duplicates are far from the target and they are unlikely to be
in the candidate set at all. In expression (\ref{eq:Poff}), the only term
that depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it
is clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
value of $\mu$ is the same for every $N$.

@@

At a $\downarrow_{/i}^-$ terminator ($i \geq 1$), the read contains a
match of length $i$ for the $+$ sequence. If the next segment is a tail,
we must make sure that the match length for the $+$ sequence does not
exceed $\gamma-1$. This means that we can have up to $\gamma-i-1$ symbols
$\square$ and thus that the weighted generating function of the tail
segments is $1 + (qz) + \ldots + (qz)^{\gamma-j-1}$.

@@

For the last two types of terminators, we must pay attention to the fact
that in general, a mismatch against $+$ that follows a mismatch against
$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
$\Downarrow_j$ if the two mismatches are within the same interval between
non-skipped positions because both sequences have match length 0. If the
mismatch against $+$ is in another interval, the terminator is
$\downarrow_{/j}^+$ because the $-$ sequence has a positive match length.

If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it cannot
be in the same interval as the preceding $\downarrow_{/i}^-$ terminator,
which imposes a lower bound on the size of the segment. Since at the
$\downarrow_{/i}^-$ terminator the match length for $+$ was $i$, there
must be a non-skipped position $i$ nucleotides before. The number of
nucleotides from the $\downarrow_{/i}^-$ terminator to the next
non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum segment
length is $y+1$ There is also an upper bound on the size of the segment
because the match length for the $+$ sequence cannot become higher than
$\gamma-1$, imposing the length to be lower than $\gamma-i$. The shortest
segment is terminated by $\downarrow_{/1}^+$, and the longest by
$\downarrow_{/{\gamma-y-i}}^+$. Finally, the weighted generating function
of segments terminated by $\downarrow_{/j}^+$ following the
$\downarrow_{/i}^-$ terminator is $(az)^{y+j-1}cz$ with $1 \leq j \leq
\gamma-y-i$. Note that for some value of $i$ and $y$, no $j$ can satisfy
the last inequality.

Finally, if the next terminator is $\Downarrow_j$ ($0 \leq j \leq n$) we
have to distinguish two cases. The first is that the segment length is $y$
nucleotides or shorter, where we have seen that a mismatch against $+$ is
sufficient to produce the $\Downarrow_j$ terminator. The second case is
that the segment length is above $y$ nucleotides, where a double mismatch
is necessary to produce the the $\Downarrow_j$ terminator. We have seen
above that the $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
before the next non-skipped position, so in the first case, only the
terminators from $\Downarrow_0$ to $\Downarrow_{y-1}$ are possible (this
case is impossible if $y=0$). More specifically, short segment terminated
by $\Downarrow_j$ have $y-j-1$ matches followed by a mismatch for the $+$
sequence, with weighted generating function $(az)^{y-j-1}bz$.

@@

As in the case of the skip encoding, a fair amount of bookkeeping is
required to specify the weighted generating functions of interest.
We reuse the quantities $a+b+c+d=1$ and give them the same meaning as in
section~\ref{sec:dual}. For notational convenience, we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big) dz, \\
  \notag
  &\quad& \text{where } x = -j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
U_{i,j}(z) &=& (az)^x
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = -i \modulo{n+1}, \\
  \notag
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& (az)^x 
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

With these definitions, one can check that the expression of the transfer
matrix $\tilde{M}_n(z)$ is

@@

The matrices $\tilde{A}(z)$ and $\tilde{C}(z)$ in the expression of
$\tilde{M}_n(z)$ are the same as in section~\ref{sec:dual}. The matrices
$\tilde{B}_n(z)$ and $\tilde{C}_n(z)$ are defined as
\begin{equation*}
\tilde{B}_n(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^+} & s_{1,1}(z) & s_{1,2}(z) &
    \ldots & s_{1,\gamma-2}(z) & s_{1,\gamma-1}(z) \\
\scriptstyle{\downarrow_{/2}^+} & s_{2,1}(z) & s_{2,2}(z) &
    \ldots & s_{2,\gamma-2}(z) & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^+} & s_{\gamma-2,1}(z) &
    s_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
\scriptstyle{\downarrow_{/\gamma-1}^+} & s_{\gamma-1,1}(z) & 0 &
    \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{equation*}
\tilde{C}_n(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^-} & t_{1,1}(z) & t_{1,2}(z) &
    \ldots & t_{1,\gamma-2}(z) & t_{1,\gamma-1}(z) \\
\scriptstyle{\downarrow_{/2}^-} & t_{2,1}(z) & t_{2,2}(z) &
    \ldots & t_{2,\gamma-2}(z) & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^-} & t_{\gamma-2,1}(z) &
    t_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
\scriptstyle{\downarrow_{/\gamma-1}^-} & t_{\gamma-1,1}(z) &
    0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
with
\begin{eqnarray*}
s_{i,j} &=&
  \begin{cases}
  cz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
  0                     &\text{ otherwise,}
  \end{cases} \\
t_{i,j} &=&
  \begin{cases}
  bz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
  0                     &\text{ otherwise,}
  \end{cases} \\
  \text{where } y &=& -i \modulo{n+1}, \text{ in both cases.} \\
\end{eqnarray*}

The computation is performed exactly as described in
section~\ref{sec:dual}. We compute the successive powers of
$\tilde{M}_n(z)$ in the arithmetic of truncated polynomials and stop the
iterations using the same criterion. So there is no change other than the
definition of the transfer matrix.

>>>>

\title{Calibrating seed-based mapping heuristics with Sesame}

@@

\begin{definition}
\label{def_dup}
The target is the DNA fragment that was actually sequenced. Duplicates are
sequences of the genome that share homology with the target (in genetics
they are often referred to as paralogs).
\end{definition}

@@

The most popular heuristic for mapping DNA sequences is a filtration
method called ``seed and extend''. The principle is to first identify
seeds, defined as short regions of high similarity between the read and
the genome, and then use an exact alignment algorithm at the seeded
locations to evaluate the candidates and identify the optimum. This
strategy was first proposed in FASTA~\cite{pmid2983426} and
BLAST~\cite{pmid2231712} to to search for homology in sequence databases
of proteins and DNA.

@@

The rest of the article is organized as follows: section~\ref{sec_seeds}
presents common seeding strategies used in bioinformatics,
section~\ref{sec:symbolic} presents the basic concepts of analytic
combinatorics that will be required to compute seeding probabilities,
sections~\ref{sec:offexact} to \ref{sec_offMEM} treat the cases of exact
seeds, skip seeds and MEM seeds, three common seeding heuristic used for
mapping, section~\ref{sec_practical} returns to the mapping problem and
revisits the assumptions of the model, section~\ref{sec:sesame} presents
Sesame, a C library implementing the main results of the theory, and
finally section~\ref{sec_disc} provides some perspectives on the present
work.

\section{Seeds}
\label{sec_seeds}

The term ``seed'' has different meanings in computational biology. It can
designate a part of the read, a part of the genome, a particular sequence
motif, or a structured pattern of matches. Also, a seed does not always
refer to an exact match. For instance, the algorithm
PatternHunter~\cite{pmid11934743} uses spaced seeds that tolerate
mismatches. To avoid any confusion, we will adopt the convention below.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}. Hash
tables are typically used to index $k$-mers, which makes them useful to
search for seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
structures have no set size so they can be used to search for seeds of
different lengths. This is the case of the FM-index, a compact data
structure based on the suffix array~\cite{manber1993suffix} and on the
Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
with a much smaller memory footprint~\cite{ferragina2000opportunistic,
ferragina2005indexing}.

@@

It is important to mention the case of spaced seeds. Originally introduced
in~\cite{califano1993flash} and popularized by
PatternHunter~\cite{pmid11934743}, spaced seeds feature ``don't-care''
positions allowing them to detect imperfect matches. Spaced seeds are
represented by models such as `11011', here indicating that the seeds have
length 5 and that the middle position is disregarded. At index time, the
genome is scanned with the model so that nucleotides labelled `1' are
concatenated and indexed. At search time, the query is scanned with the
model and the concatenated nucleotides labelled `1' are looked up in the
index.

Spaced seeds of weight $w$ (the number of 1s in the model) have the same
memory requirements as continuous seeds of length $w$ but they are more
sensitive~\cite{li2006superiority}, making them very attractive for
homology search. They were used in the first generation of short-read
mappers~\cite{pmid18684737, pmid19461883, pmid19675096, pmid2945724}, but
nowadays they find more applications in genome
comparisons~\cite{pmid21209072, healy2016flak},
metagenomics~\cite{pmid26209798, pmid27540266}, genome
assembly~\cite{pmid26539459} and long-read mapping~\cite{pmid27079541}.

The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
supports continuous seeds. Some workarounds are
available~\cite{horton2008dislex, gagie2017compressed} but they increase
the memory footprint, explaining that short reads are typically mapped
using continuous seeds. More generally, computing the sensitivity of
spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
martin2017faster}. It is theoreticlally possible to do this using the
tools introduced below, but the complexity becomes prohibitive for seed
models with several don't-care positions. For these reasons we will not
treat the case of spaced seeds and we will give more detail in
section~\ref{sec_disc} bout the kind of difficulties they represent for
our approach.

@@

\begin{definition}
A Maximal Exact Match (MEM) is a subsequence of the read that is present
in the reference genome and that cannot be extended---either because the
read ends or because the extended subsequence is not in the genome.
\end{definition}

@@

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Figure~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching
the target and the first duplicate (provided $\gamma \leq 12$).

@@

We need to know whether a nucleotide is a sequencing error, but this time
we also need to know its phase in the repeated cycles of skipped
positions. For this, we define the skip-$n$ alphabet $\mathcal{A}_n =
\{\square, |, \Downarrow_0, \Downarrow_1, \ldots, \Downarrow_n \}$. Again,
the symbol $\square$ represents a correct nucleotide and the symbol $|$ is
a terminator added at the end of the read. The symbols $\Downarrow_j$ ($ 0
\leq j \leq n$) represent sequencing errors and $j$ indicates the number
of nucleotides until the next non-skipped position (\textit{i.e.} $j=0$
for nucleotides immediately before a non-skipped position and $j=n$ for
nucleotides at a non-skipped position).

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the $\square$ symbol
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$j$ in $\Downarrow_j$ indicates the number of nucleotidesu until the next
non-skipped position. For $j=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

Those segments follow a $\Downarrow_i$-segment, so they start $i$
nucleotides before the next non-skipped position. The shortest following
segment that ends $j$ nucleotides before a non-skipped position has length
$\ell_0 = i-j$ if $i > j$, and $n+1-j+i$ otherwise. This is equivalent to
defining $\ell_0$ as $x+1$ where $x$ is the number of $\square$ symbols of
the shortest $\Downarrow_j$-segment, \textit{i.e.} $x = i-j-1
\modulo{n+1}$.

@@

\subsection{Transfer matrix}
\label{sec_transmat_skip}

@@

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, the top right entry of
$\tilde{M}_0(z)^{s+1}$ is the weighted generating function of reads that
contain $s$ mismatches (where doule mismatches count as one). We thus
define $\tilde{p}$ as the upper bound on the probability of a mismatch,
\textit{i.e.} $\tilde{p} = \max\{b,c,d\}$. The updated formula
(\ref{eq:bound}) now gives an upper bound of the probability that a read
of size $k$ contains $s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.070. When $\mu$ is much smaller, the duplicates tend to
be exactly identical to the target, making it impossible that there is a
seed for a duplicate but no seed for the target. When $\mu$ is much
larger, the duplicates are far from the target and they are unlikely to be
in the candidate set at all. In expression (\ref{eq:Poff}), the only term
that depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it
is clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
value of $\mu$ is the same for every $N$.

@@

At a $\downarrow_{/i}^-$ terminator ($i \geq 1$), the read contains a
match of length $i$ for the $+$ sequence. If the next segment is a tail,
we must make sure that the match length for the $+$ sequence does not
exceed $\gamma-1$. This means that we can have up to $\gamma-i-1$ symbols
$\square$ and thus that the weighted generating function of the tail
segments is $1 + (qz) + \ldots + (qz)^{\gamma-i-1}$.

@@

For the last two types of terminators, we must pay attention to the fact
that in general, a mismatch against $+$ that follows a mismatch against
$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
$\Downarrow_j$ if the two mismatches are within the same interval between
non-skipped positions because both sequences lovally have match length 0.
If the mismatch against $+$ is in another interval, the terminator is
$\downarrow_{/j}^+$ because in that case the $-$ sequence has a positive
match length. For now, bear in mind that a mismatch against the $+$
sequence only can produce a $\Downarrow_j$ terminator ($0 \leq j \leq n$),
this will be important later.

If the next terminator is $\downarrow_{/j}^+$ ($j \geq 1$), then it must
be separated from the  preceding $\downarrow_{/i}^-$ terminator by a
non-skipped position, which imposes a lower bound on the size of the
segment. Since at the $\downarrow_{/i}^-$ terminator the match length for
$+$ was $i$, there must be a non-skipped position $i$ nucleotides before.
The number of nucleotides from the $\downarrow_{/i}^-$ terminator to the
next non-skipped position is thus $y = -i \modulo{n+1}$, so the minimum
segment length is $y+1$ There is also an upper bound on the size of the
segment because the match length for the $+$ sequence cannot become higher
than $\gamma-1$, imposing the length to be lower than $\gamma-i$. The
shortest segment is terminated by $\downarrow_{/1}^+$, and the longest by
$\downarrow_{/{\gamma-y-i}}^+$. Finally, the weighted generating function
of segments terminated by $\downarrow_{/j}^+$ following the
$\downarrow_{/i}^-$ terminator is $(az)^{y+j-1}cz$ with $1 \leq j \leq
\gamma-y-i$. Note that for some value of $i$ and $y$, no $j$ can satisfy
the last inequality.

Finally, if the next terminator is $\Downarrow_j$ ($0 \leq j \leq n$) we
have to distinguish two cases, depending on whether the segment ends with
a double mismatch, or with a single mismatch against the $+$ sequence. For
the first case we can apply the rationale of
section~\ref{sec:example_skip}. Recall that the index $j$ indicates the
number of nucleotides until the next non-skipped position. Let $\ell_0$ be
the length of the shortest possible segment. The lengths of the other
segments are of the form $\ell_0 + n+1, \ldots, \ell_0 + m(n+1)$, where
the integer $m$ must be chosen so that the segment does not create a seed.
At the $\downarrow_{/i}^-$ terminator, there is a match of length $i$ for
the $+$ sequence so $m$ is the largest integer such that $i+\ell_0+m(n+1)
< \gamma$, \textit{i.e.} $m = \lfloor(\gamma-i-\ell_0) / (n+1)\rfloor$.

The $\downarrow_{/i}^-$ terminator is $y = -i \modulo{n+1}$ nucleotides
before the next non-skipped position so the shortest segment has length
$\ell_0 = y-j$ if $y > j$, and $n+1-j+y$ otherwise. This is equivalent to
defining $\ell_0$ as $x+1$ where $x$ is the number of $\square$ symbols in
the shortest segment, \textit{i.e.} $x=-i-j-1 \modulo{n+1}$. Summing the
weighted generating functions of the individual segments, we find
$(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz$.

The last remaining issue is that a mismatch against the $+$ sequence only
can produce a $\Downarrow_j$ terminator. This happens when the preceding
$\downarrow_{/i}^-$ terminator is not separated from the mismatch by a
non-skipped position (in this case both sequences locally have match
length 0). The $\downarrow_{/i}^-$ terminator is located $y$ nucleotides
before the next non-skipped position, so if the terminator is from
$\Downarrow_0$ to $\Downarrow_{y-1}$ we need to add the term
$(az)^xbz$ to the previous weighted generating function. In
conclusion, the weighted generating function of segments terminated by
$\Downarrow_j$ is $(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz +
\delta^+_{i,j}(z)$, where $\delta^+_{i,j}(z) = (az)^xbz$ if $j < y$ and
$0$ otherwise.

We can find the weighted generating functions by just reversing the $+$
and $-$ signs in the previous section. This way we can see that the
weighted generating function of tail segments is $1 + (qz) + \ldots +
(qz)^{\gamma-i-1}$.

Likewise, the weighted generating function of segments terminated by
$\downarrow_{/j}^+$ is $(qz)^{i-j-1}bz$ with $i < j \leq \gamma-1$.

The weighted generating function of segments terminated by
$\downarrow_{/j}^-$ is $(az)^{y+j-1}dz$ with $y = -i \modulo{n+1}$ and $1
\leq j \leq \gamma-y-i$.

Finally, the weighted generating function of segments terminated by
$\Downarrow_j$ is $(az)^x(1+(az)^{n+1} + \ldots + (az)^{m(n+1)})dz +
\delta^-_{i,j}(z)$, where $\delta^-_{i,j}(z) = (az)^xcz$ if $j < y$ and
$0$ otherwise.

@@

We now have all the elements to specify the transfer matrix of reads with
no skip-$n$ seed for either sequence. Recall that $n$ is the number of
skipped nucleotides, $\gamma$ is the minimum seed length, $a$ is the
probabiliy of a double match, $b$ is the probability of a mismatch against
the $-$ sequence only, $c$ is the probability of a mismatch against the
$+$ sequence only and $d$ is the probability of a double mismatch. For
notational convenience we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big) dz, \\
  \notag
  &\quad& \text{where } x = -j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
U_{i,j}(z) &=& (az)^x
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = -i \modulo{n+1}, \\
  \notag
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& (az)^x 
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

With these notations, the information from the previous sections can be
summarized in the transfer matrix $\tilde{M}_n(z)$ equal to

@@

The matrices $\tilde{A}(z)$ and $\tilde{C}(z)$ in the expression of
$\tilde{M}_n(z)$ are the same as in section~\ref{sec:dual}. They are
reproduced here for convenience.
\begin{equation*}
\tilde{A}(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^+} & 0 & r_0^+(z) & \ldots &
    r_{\gamma-4}^+(z) & r_{\gamma-3}^+(z) \\
\scriptstyle{\downarrow_{/2}^+} & 0 & 0 & \ldots &
    r_{\gamma-5}^+(z) & r_{\gamma-4}^+(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^+} & 0 & 0 & \ldots & 0 & r_0^+(z) \\
\scriptstyle{\downarrow_{/\gamma-1}^+} & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{equation*}
\tilde{C}_0(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^-} & r_0^+(z) & r_1^+(z) & \ldots &
    r_{\gamma-3}^+(z) & r_{\gamma-2}^+(z) \\
\scriptstyle{\downarrow_{/2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
    r_{\gamma-3}^+(z) & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^-} & r_0^+(z) & r_1^+(z) & \ldots &
    0 & 0 \\
\scriptstyle{\downarrow_{/\gamma-1}^-} & r_0^+(z) & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

The matrices $\tilde{B}_n(z)$ and $\tilde{C}_n(z)$ are defined as
\begin{equation*}
\tilde{B}_n(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^-} & \scriptstyle{\downarrow_{/2}^-} &
   \ldots & \scriptstyle{\downarrow_{/\gamma-2}^-} &
   \scriptstyle{\downarrow_{/\gamma-1}^-} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^+} & s_{1,1}(z) & s_{1,2}(z) &
    \ldots & s_{1,\gamma-2}(z) & s_{1,\gamma-1}(z) \\
\scriptstyle{\downarrow_{/2}^+} & s_{2,1}(z) & s_{2,2}(z) &
    \ldots & s_{2,\gamma-2}(z) & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^+} & s_{\gamma-2,1}(z) &
    s_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
\scriptstyle{\downarrow_{/\gamma-1}^+} & s_{\gamma-1,1}(z) & 0 &
    \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{equation*}
\tilde{C}_n(z) = 
\begin{blockarray}{cccccc}
   & \scriptstyle{\downarrow_{/1}^+} & \scriptstyle{\downarrow_{/2}^+} &
    \ldots & \scriptstyle{\downarrow_{/\gamma-2}^+} &
    \scriptstyle{\downarrow_{/\gamma-1}^+} \\
\begin{block}{c[ccccc]}
\scriptstyle{\downarrow_{/1}^-} & t_{1,1}(z) & t_{1,2}(z) &
    \ldots & t_{1,\gamma-2}(z) & t_{1,\gamma-1}(z) \\
\scriptstyle{\downarrow_{/2}^-} & t_{2,1}(z) & t_{2,2}(z) &
    \ldots & t_{2,\gamma-2}(z) & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\scriptstyle{\downarrow_{/\gamma-2}^-} & t_{\gamma-2,1}(z) &
    t_{\gamma-2,2}(z) & \ldots & 0 & 0 \\
\scriptstyle{\downarrow_{/\gamma-1}^-} & t_{\gamma-1,1}(z) &
    0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
with
\begin{eqnarray*}
s_{i,j} &=&
  \begin{cases}
  cz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
  0                     &\text{ otherwise,}
  \end{cases} \\
t_{i,j} &=&
  \begin{cases}
  bz \cdot (az)^{y+j-1} &\text{ if } i+j+y \leq \gamma \\
  0                     &\text{ otherwise,}
  \end{cases} \\
  \text{where } y &=& -i \modulo{n+1}, \text{ in both cases.} \\
\end{eqnarray*}
